---
author: "Melody Jin"
title: "cellXY"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('cellXY')`"
vignette: >
    %\VignetteEncoding{UTF-8}
    %\VignetteIndexEntry{cellXY}
    %\VignetteEngine{knitr::rmarkdown}
output: >
    BiocStyle::html_document
html_document:
    fig_caption: yes
    fig_retina: FALSE
    keep_md: FALSE
editor_options: 
    chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cellXY)
library(SingleCellExperiment)
library(ggplot2)
library(scRNAseq)
library(reshape2)
```

We will use MarquesBrainData from the scRNAseq package to illustrate how to use functions in cellXY. There are 5069 cells from 4 samples. We can see that 1060 cells are Female, 1560 cells are M, and the sex information for the remaining cells are unknown.

# Load the data
```{r}
sce <- MarquesBrainData()

table(sce$Sex)
```

We can find Male/Female doublets by calling the `findMFDoublet` function.

# Find Male/Female doublets

```{r}
db_pred =findMfDoublet(x=counts(sce),genome="Mm")
db_pred$sex = sce$Sex
print(table(db_pred$prediction, db_pred$sex))
```

We identify 485 Male/Female doublets in this dataset. Most of the selected doublets do not have a known sex label. If you want to see the distribution of features used for finding those doublets, you can call `preprocessDb` which will give you a list with two data frames. Each data frame will have rows as cells and columns as features. One data frame contains the raw count values, and the other data frame contains the normalised count.

```{r}

prep_db_lst = preprocessDb(x=counts(sce),genome="Mm",qc = FALSE)
# normalised features
df_ps = prep_db_lst$data.df
# raw count of features 
tcm_ps = prep_db_lst$tcm.final

vis_df = df_ps
colnames(df_ps)[4]="libsize"
df_ps$prediction = db_pred$prediction
df_ps$sample = db_pred$sex
vis_df = reshape2::melt(df_ps, id.vars =c("prediction","sample"), variable.name ="type")
vis_df$prediction = factor(vis_df$prediction)
vis_df$sex = factor(vis_df$sample)
vis_df$type = factor(vis_df$type)


ggplot(data = vis_df, aes(x=type, y=value, fill=prediction)) +
  facet_wrap(~sample, ncol=4)+
  #geom_violin(scale="width")+
  geom_boxplot(outlier.shape = NA)+
  geom_point(position=position_jitterdodge(jitter.width=0.3, jitter.height=0), 
             size=0.05,  alpha=0.9)+
  scale_fill_manual(values=c("#BC80BD","#4DAF4A"), name="")+

  theme_bw()+
    theme(legend.title=element_text(size=10),
        axis.title.x=element_text(size=10),
        axis.title.y=element_text(size=10), 
        panel.spacing = unit(0.3, "lines"), 
        legend.position="top",
        legend.text=element_text(margin=ggplot2::margin(r=0.3,unit="cm"), size=18),
        strip.text = element_text(size = rel(1.3)))+
  xlab("")+ylab(expression(log[2]~"normalised count"))

```

Then we can predict the sex labels for those singlets by calling `classifySex` function. 
# Predict sex labels for singlets
```{r}

sce_s = colnames(counts(sce))[db_pred$prediction=="Singlet"]

sc_pred=classifySex(x=counts(sce)[, sce_s],genome="Mm", qc=TRUE)
table(sc_pred$prediction)

sc_pred$sex = sce$Sex[match(sce_s, colnames(sce))]
table(sc_pred)
```

We identify 1735 female cells and 2813 male cells. We can not classify one cell and pick another 17 cells as low-quality cells. Thus there are 18 overall cells that are labelled as NA. If you want to see the distribution of features used for predicting the sex labels, you can call `preprocess`, which will give you a list with two data frames. Each data frame will have rows as cells and columns as features. One data frame contains the raw count values, and the other data frame contains the normalised count.

```{r}

prep_sc_lst = preprocess(x=counts(sce)[, sce_s],genome="Mm",qc = FALSE)
# normalised features
df_ps = prep_sc_lst$data.df[sce_s,]
# raw count of features 
tcm_ps = prep_sc_lst$tcm.final[sce_s,]

vis_df = df_ps
df_ps$prediction = sc_pred$prediction
df_ps$sample = sc_pred$sex
vis_df = reshape2::melt(df_ps, id.vars =c("prediction","sample"), variable.name ="type")
vis_df$prediction = factor(vis_df$prediction)
vis_df$sex = factor(vis_df$sample)
vis_df$type = factor(vis_df$type)

ggplot(data = vis_df[vis_df$prediction!="NA", ], aes(x=type, y=value, fill=prediction)) +
  facet_wrap(~sample, ncol=4)+
  geom_boxplot(outlier.shape = NA)+
  geom_point(position=position_jitterdodge(jitter.width=0.6, jitter.height=0), 
             size=0.1,  alpha=0.9)+
  scale_fill_manual(values=c("#FFB900FF","#5773CCFF"), name="")+
  theme_bw()+
  theme(legend.title=element_text(size=10),
        axis.title.x=element_text(size=10),
        axis.title.y=element_text(size=10), 
        panel.spacing = unit(0.3, "lines"), 
        legend.position="top",
        legend.text=element_text(margin=ggplot2::margin(r=0.3,unit="cm"), size=18),
        strip.text = element_text(size = rel(1.3)))+
  xlab("")+ylab(expression(log[2]~"normalised count"))

```
